The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2016-11-29 manuscript/acronyms.tex
add gsea acronym
use long names for genes where possible

\DeclareAcronym{EV}{short=EV,long=Extracellular Vesicle}
\DeclareAcronym{SVM}{long=Support Vector Machine}
\DeclareAcronym{ROC}{long=Receiver Operating Characteristic}
\DeclareAcronym{KNN}{long=K Nearest Neighbors}
\DeclareAcronym{SRA}{long=Sequence Read Archive}
\DeclareAcronym{PRG2}{short=\textit{PRG2},long=Proteoglycan 2, bone marrow}
\DeclareAcronym{FPKM}{long=Fragments Per Kilobase of exon per Million reads}
\DeclareAcronym{CSH1}{short=\textit{CSH1},long=chorionic somatomammotropin hormone 1}
\DeclareAcronym{CSH2}{short=\textit{CSH2},long=chorionic somatomammotropin hormone 2}
\DeclareAcronym{CSHL1}{short=\textit{CSHL1},long=chorionic somatomammotropin hormone-like 1}
\DeclareAcronym{GH2}{short=\textit{GH2},long=growth hormone 2}
\DeclareAcronym{CEA}{long=carcinoembryonic antigen}
\DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long=3-β Hydroxysteroid dehydrogenase 1}
\DeclareAcronym{GSE}{long=Gene Set Enrichment}
\DeclareAcronym{OU}{long=Ornstein-Uhlenbeck}
\DeclareAcronym{OUCH}{long=\ac{OU} Comparative Hypotheses}
\DeclareAcronym{TSI}{long=Tissue Specificity Index}
\DeclareAcronym{SRA}{long=Sequence Read Archive}
\DeclareAcronym{GH}{long=Growth Hormone}
\DeclareAcronym{PSG}{long=Pregnancy-Specific Glycoprotein}
\DeclareAcronym{PAG}{long=Pregnancy-Associated Glycoprotein}
\DeclareAcronym{GO}{long=Gene Ontology}
\DeclareAcronym{FDR}{long=False Discovery Rate}

>>>>

\DeclareAcronym{EV}{short=EV,long=Extracellular Vesicle}
\DeclareAcronym{SVM}{long=Support Vector Machine}
\DeclareAcronym{ROC}{long=Receiver Operating Characteristic}
\DeclareAcronym{KNN}{long=K Nearest Neighbors}
\DeclareAcronym{SRA}{long=Sequence Read Archive}
\DeclareAcronym{PRG2}{short=\textit{PRG2},long=Proteoglycan 2, bone marrow}
\DeclareAcronym{FPKM}{long=Fragments Per Kilobase of exon per Million reads}
\DeclareAcronym{CSH1}{short=\textit{CSH1},long=chorionic somatomammotropin hormone 1}
\DeclareAcronym{CSH2}{short=\textit{CSH2},long=chorionic somatomammotropin hormone 2}
\DeclareAcronym{CSHL1}{short=\textit{CSHL1},long=chorionic somatomammotropin hormone-like 1}
\DeclareAcronym{GH2}{short=\textit{GH2},long=growth hormone 2}
\DeclareAcronym{CEA}{long=carcinoembryonic antigen}
\DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long=3-β Hydroxysteroid dehydrogenase 1}
\DeclareAcronym{GSE}{long=Gene Set Enrichment}
\DeclareAcronym{OU}{long=Ornstein-Uhlenbeck}
\DeclareAcronym{OUCH}{long=\ac{OU} Comparative Hypotheses}
\DeclareAcronym{TSI}{long=Tissue Specificity Index}
\DeclareAcronym{SRA}{long=Sequence Read Archive}
\DeclareAcronym{GH}{long=Growth Hormone}
\DeclareAcronym{PSG}{long=Pregnancy-Specific Glycoprotein}
\DeclareAcronym{PAG}{long=Pregnancy-Associated Glycoprotein}
\DeclareAcronym{GO}{long=Gene Ontology}
\DeclareAcronym{FDR}{long=False Discovery Rate}
\DeclareAcronym{gsea}{short=GSEA,long=gene set enrichment analysis}
\DeclareAcronym{CRH}{short=\textit{CRH},long={corticotropin-releasing hormone}}
\DeclareAcronym{KISS1}{short=\textit{KISS1},long={KiSS-1 metastatis-supressor}}
\DeclareAcronym{PAPPA}{short=\textit{PAPPA},long={pappalysin 1}}
\DeclareAcronym{ADAM12}{short=\textit{ADAM12},long={ADAM Metallopeptidase Domain 12}}
\DeclareAcronym{ANXA1}{short=\textit{ANXA1},long={annexin A1}}
\DeclareAcronym{ANXA5}{short=\textit{ANXA5},long={annexin A5}}
\DeclareAcronym{S100A11}{short=\textit{S100A11},long={S100 calcium binding protein A11}}
\DeclareAcronym{S100A10}{short=\textit{S100A10},long={S100 calcium binding protein A10}}
\DeclareAcronym{LGALS1}{short=\textit{LGALS1},long={galectin 1}}
\DeclareAcronym{LAMC1}{short=\textit{LAMC1},long={laminin subunit gamma 1}}
\DeclareAcronym{LUM}{short=\textit{LUM},long={lumican}}
\DeclareAcronym{GRB2}{short=\textit{GRB2},long={growth factor receptor bound protein 2}}
\DeclareAcronym{RALB}{short=\textit{RALB},long={RAS like proto-oncogene B}}
\DeclareAcronym{ERVFRD-1}{short=\textit{ERVFRD-1},long={endogenous retrovirus group FRD member 1}}
\DeclareAcronym{ERVW-1}{short=\textit{ERVW-1},long={endogenous retrovirus group W member 1}}
\DeclareAcronym{MFSD2A}{short=\textit{MFSD2A},long={major facilitator superfamily domain containing 2A}}
\DeclareAcronym{PRSS16}{short=\textit{PRSS16},long={protease, serine 16}}
\DeclareAcronym{TIMP3}{short=\textit{TIMP3},long={TIMP metallopeptidase inhibitor 3}}
\DeclareAcronym{F11R}{short=\textit{F11R},long={F11 receptor}}
\DeclareAcronym{SDC4}{short=\textit{SDC4},long={syndecan 4}}
\DeclareAcronym{FN1}{short=\textit{FN1},long={fibronectin 1}}
\DeclareAcronym{ITGB1}{short=\textit{ITGB1},long={integrin subunit beta 1}}
\DeclareAcronym{ITGB1BP1}{short=\textit{ITGB1BP1},long={integrin subunit beta 1 binding protein 1}}
\DeclareAcronym{EREG}{short=\textit{EREG},long={epiregulin}}
\DeclareAcronym{UNC93A}{short=\textit{UNC93A},long={unc-93 homolog A (\textit{C. elegans})}}
\DeclareAcronym{UNC93B1}{short=\textit{UNC93B1},long={unc-93 homolog B1 (\textit{C. elegans)}}}
\DeclareAcronym{WNT5A}{short=\textit{WNT5A},long={Wnt family member 5A}}
\DeclareAcronym{IL1RL1}{short=\textit{IL1RL1},long={interleukin 1 receptor like 1}}
\DeclareAcronym{PLAC1}{short=\textit{PLAC1},long={placenta specific 1}}
\DeclareAcronym{Afp}{short=\textit{Afp},long={alpha fetoprotein}}
\DeclareAcronym{Gjb2}{short=\textit{Gjb2},long={gap junction protein beta 2}}
\DeclareAcronym{CD9}{short=\textit{CD9},long={CD9 molecule}}

### 2016-11-29 manuscript/manuscript.Rnw
use long names for genes where possible

  \textbf{Results} We identified a set of
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  core genes which are expressed (FPKM $\ge 10$) in all placentas, %
  %
  including genes with immune-modulating properties (\textit{ANXA2},
  \textit{ANXA1}, \textit{S100A11}, \textit{S100A10}, and \textit{LGALS1}), cell-cell
  interactions (\textit{LAMC1}, \textit{LUM}, and \textit{LGALS1}),
  invasion (\textit{GRB2} and \textit{RALB5}) and syncytialization
  (\textit{ANXA5} and \textit{ANXA1}). %
  We also identified multiple pre-eclampsia associated genes which are
  differentially expressed in \textit{Homo sapiens} when compared to
  the other %
  \Sexpr{star.logs[,length(unique(species))-1]} species. %

  \textbf{Discussion} %
  Using the strictest filtering strategy we find a core set of only
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  genes is essential for the proper functioning of the placenta in all
  species analyzed. We also found significantly higher expression of
  the pre-eclampsia associated genes \textit{KISS1}, \textit{CRH},
  \textit{PAPPA}, and \textit{ADAM12} on the \human lineage.

@@

@@

Previous studies of some species (i.e., \human, \mouse, \cow, and
\elephant) have provided some preliminary answers to these questions.
They revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}}.
For example, each superorder has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
independent expansions of the prolactins in \mouse, \carmeli, \galili,
and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
related to the thymus-specific serine protease \textit{PRSS16} in
\elephant (in this work), and the expansion of the growth hormone and
chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

@@

\DEW{Add gene names}

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \human lineage in comparison to all
other lineages (\cref{tab:human_lineage_specific}, Supplemental
\cref{tab:human_lineage_specific_all}). These genes include multiple
genes which are thought to be involved in pre-eclampsia. Notably,
\textit{CRH} (FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
\cref{fig:crh_evolution_tree}), has been shown to be more highly
expressed in placentas from pre-eclamptic pregnancies
\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
where it is hypothesized to activate decidual macrophages. Likewise,
\textit{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
\cref{fig:kiss1_evolution_tree}) also shows increased expression in
pre-eclamptic placentas
\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
involved in modulating trophoblast invasion. Two other genes,
\textit{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
\cref{fig:pappa_evolution_tree}) and \textit{ADAM12}
(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}) are also differentially expressed
but are negatively correlated with pre-eclampsia
\cite{Spencer.ea2008:ADAM12sinmaternalserum}.

@@

Multiple components of annexin complexes are included in the core
transcriptome, including \textit{ANXA2}, \textit{ANXA1},
\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\textit{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.

Another annexin which is highly expressed in all term placentas is
\textit{ANXA5}, which was originally identified as a an anti-coagulant
protein, but is also involved in syncitium formation, membrane repair,
and reduction of spontaneous
abortion~\cite{Ueki.ea2012:Lossofmaternalannexin}. Other notable
syncytiotrophoblast fusion proteins, Syncytin-1 (\textit{ERVW-1}) and -2
(\textit{ERVFRD-1}) were not present in the core placenta, as they (or
their homologs) are not highly expressed in \dog, \bonobo, or \sheep,
nor was \textit{MFSD2A}, the Syncytin-2 receptor (\ac{FPKM} $\ge 10$
only in \human, \mouse, and \dog).

@@

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome, translation, signal
recognition particle targeting to the rough endoplasamic reticulum,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway and the Actin Y
pathway. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \textit{FN1} and \textit{ITG}
are highly expressed, which activate Rac, and potentially lead to the
formation of focal adhesions and lamellipodia formation through their
inhibition of GSN and HSPC300. Other components of cell-cell
adhesion, including \textit{F11R} which forms tight junctions in
epithelia, proteglycans \textit{SDC4} and \textit{LUM}, and the
laminin \textit{LAMC1}.

@@

Supplemental \cref{tab:shape_genes} shows the 50 genes which are most
significantly associated with the shape of placenta attachment, as
well as the shape in which the difference is most significant. The
most significant genes which were associated with diffuse placenta
attachment include \textit{EREG}, a ligand of EGFR, \textit{UNC93A}, a
homolog of \textit{UNC93B1} which is involved in the balancing between
Tlr7 and Tlr9 immune response in
mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
\textit{WNT5A}, which is a member of the Wnt ligand family whose
singaling pathway has been implicated in motility in human trophoblast
cells~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}.

@@

The identification of multiple genes (\textit{CRH}, \textit{ADAM12},
\textit{KISS1}, \textit{PAPPA}, \textit{IL1RL1}, and \textit{PLAC1};
\cref{tab:human_lineage_specific}) which are associated with
pre-eclampsia which have increased expression on the \human lineage
may explain the larger proportion of pre-eclampsia observed in humans.
However, there are still genes which are associated with pre-eclampsia
which are also more highly expressed in the \mouse lineage such as
\textit{Afp} and \textit{Gjb2}. Furthermore, the onset of
pre-eclampsia is well before term; all of our placentas are term
placentas and provide limited information about the transcriptome
during the onset of pre-eclampsia.

@@

\subsubsection{Immune Regulation and Annexins}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation.

@@

While Syncytin-1 and -2 are implicated in the formation of the
syncytiotrophoblast in
\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
of \textit{ERVFRD-1} and \textit{ERVW-1} are widely expressed in most
of the placentas analyzed here, they are not expressed in \bonobo,
\sheep, and \dog, even though these mammals form
syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
Because the formation of syncytiotrophoblast is the stem state of
Eutheria, Afrotheria, Euarchontoglires, and
Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
the gene(s) responsible for cytotrophoblast fusion to form
syncytiotrophoblast is likely to be highly expressed in the core
placenta. \textit{ANXA5} is known to be expressed at multiple stages
of placentation, and is capable of repairing membrane defects and
promoting the fusion of
cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
the presence of Ca$^{2+}$. While there are likely other genes which
are responsible for the positioning of cytotrophoblast cells in close
proximity to start the fusion process, \textit{ANXA5} is likely to be
essential for the completion of fusion and the maintenance of membrane
repair during the massive flux of nutrients and waste in the placenta.

@@

The high expression of components of the EGFR signaling pathway in the
core placenta transcriptome (including \textit{GRB2}, \textit{RALB})
as well as annexins which are involved in the location of EGF receptor
localization (\textit{ANXA1} and \textit{ANXA2}) in the core placenta
is consistent with previous findings showing that alterations to the
EGF signaling system change invasion of trophoblast cells into the
spiral arteries~\cite{Armant.ea2015:Reducedexpressionofepidermal}.
Another component of the core placenta transcriptome, \textit{CD9},
has been shown to suppress \human extravillous trophoblast
invasion~\cite{Matsumoto.ea2016:CD9suppresseshumanextravillous}, and
we hypothesize that it is responsible for maintaining the balance
between invasion and attachment in the other placentas examined here
as well.

@@

The core placenta transcriptome contains genes which are involved in
maternal-fetal tolerance, such as the annexins, and genes responsible
for cell-cell and cell-matrix interactions, such as \textit{LGALS1}.
These genes were almost certainly present in the placenta of the
common mammalian ancestor. We hypothesize that modifications to their
expression or function will compromise the development and
organization of the placenta.

We also identified genes differentially expressed on the \human
lineage which are implicated in pre-eclampsia, including
\textit{ADAM12}, \textit{CRH}, and \textit{KISS1}.

>>>>

  \textbf{Results} We identified a set of
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  core genes which are expressed (FPKM $\ge 10$) in all placentas, %
  %
  including genes with immune-modulating properties (\ac{ANXA2},
  \ac{ANXA1}, \ac{S100A11}, \ac{S100A10}, and \ac{LGALS1}), cell-cell
  interactions (\ac{LAMC1}, \ac{LUM}, and \ac{LGALS1}),
  invasion (\ac{GRB2} and \ac{RALB}) and syncytialization
  (\ac{ANXA5} and \ac{ANXA1}). %
  We also identified multiple pre-eclampsia associated genes which are
  differentially expressed in \textit{Homo sapiens} when compared to
  the other %
  \Sexpr{star.logs[,length(unique(species))-1]} species. %

  \textbf{Discussion} %
  Using the strictest filtering strategy we find a core set of only
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  genes is essential for the proper functioning of the placenta in all
  species analyzed. We also found significantly higher expression of
  the pre-eclampsia associated genes \ac{KISS1}, \ac{CRH},
  \ac{PAPPA}, and \ac{ADAM12} on the \human lineage.

@@

% Reset all acronyms
\acresetall

@@

Previous studies of some species (i.e., \human, \mouse, \cow, and
\elephant) have provided some preliminary answers to these questions.
They revealed a stunning divergence in the genetics of the placenta,
especially in later stages of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}}.
For example, each superorder has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
independent expansions of the prolactins in \mouse, \carmeli, \galili,
and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
related to the thymus-specific serine protease \ac{PRSS16} in
\elephant (in this work), and the expansion of the growth hormone and
chorionic gonadotropin beta genes in anthropoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.

@@

% \DEW{Add gene names}

Using a mixed-effects linear model to account for repeated testing of
\human and \mouse, we identified multiple genes which are
differentially expressed in the \human lineage in comparison to all
other lineages (\cref{tab:human_lineage_specific}, Supplemental
\cref{tab:human_lineage_specific_all}). These genes include multiple
genes which are thought to be involved in pre-eclampsia. Notably,
\ac{CRH} (FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
\cref{fig:crh_evolution_tree}), has been shown to be more highly
expressed in placentas from pre-eclamptic pregnancies
\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
where it is hypothesized to activate decidual macrophages. Likewise,
\ac{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
\cref{fig:kiss1_evolution_tree}) also shows increased expression in
pre-eclamptic placentas
\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
involved in modulating trophoblast invasion. Two other genes,
\ac{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
\cref{fig:pappa_evolution_tree}) and \ac{ADAM12}
(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
\cref{fig:adam12_evolution_tree}) are also differentially expressed
but are negatively correlated with pre-eclampsia
\cite{Spencer.ea2008:ADAM12sinmaternalserum}.

@@

Multiple components of annexin complexes are included in the core
transcriptome, including \ac{ANXA2}, \ac{ANXA1},
\ac{S100A11}, \ac{S100A10}. ANXA1 is involved in the
resolution of the inflammation response through inhibition of
phospholipase A2 activity and signaling through the formyl peptide
receptor family, and has previously been hypothesized to be important
for the maintenance of the anti-inflammatory state during pregnancy
\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
known to interact with S100A10, S100A6, S100A11, and S100A4
\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
its interaction with S100P as well; it is involved in cell-cell
interaction as well as vesicle trafficking and von Willibrand Factor
secretion. The expression regulation of another core gene,
\ac{TIMP3}, has been implicated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.

Another annexin which is highly expressed in all term placentas is
\ac{ANXA5}, which was originally identified as a an anti-coagulant
protein, but is also involved in syncitium formation, membrane repair,
and reduction of spontaneous
abortion~\cite{Ueki.ea2012:Lossofmaternalannexin}. Other notable
syncytiotrophoblast fusion proteins, Syncytin-1 (\ac{ERVW-1}) and -2
(\ac{ERVFRD-1}) were not present in the core placenta, as they (or
their homologs) are not highly expressed in \dog, \bonobo, or \sheep,
nor was \ac{MFSD2A}, the Syncytin-2 receptor (\ac{FPKM} $\ge 10$
only in \human, \mouse, and \dog).

@@

The core placenta transcriptome is significantly enriched in genes
which are involved in multiple biological pathways, including multiple
pathways related to translation (ribosome, translation, signal
recognition particle targeting to the rough endoplasamic reticulum,
etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
translation, the next most abundant set of significantly enriched
molecular pathways are related to the migration of the trophoblast
into the endometrium, including the EGFR1 pathway and the Actin Y
pathway. Two other significant pathways, the mPR/Progesterone pathway
and Rho cell motility pathways, feed into the Actin Y pathway as well.
Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
humans with the median expression in FPKM of genes in the pathway
shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
expressed; their activation leads to an increase in migration and
invasion in response to EGF binding to
EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
cytoskeleton pathway is depicted in Supplemental
\cref{fig:kegg_actin}, which shows that \ac{ITGB1} and its binding
protein \ac{ITGB1BP1} as well as \ac{FN1} are highly expressed, which
activates Rac. This potentially leads to the formation of focal adhesions
and lamellipodia formation through inhibition of GSN and
HSPC300. Other components of cell-cell adhesion, including \ac{F11R}
which forms tight junctions in epithelia, the proteglycans \ac{SDC4} and
\ac{LUM}, and \ac{LAMC1}.

@@

Supplemental \cref{tab:shape_genes} shows the 50 genes which are most
significantly associated with the shape of placenta attachment, as
well as the shape in which the difference is most significant. The
most significant genes which were associated with diffuse placenta
attachment include \ac{EREG}, a ligand of EGFR, \ac{UNC93A}, a
homolog of \ac{UNC93B1} which is involved in the balancing between
Tlr7 and Tlr9 immune response in
mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
\ac{WNT5A}, which is a member of the Wnt ligand family whose
signaling pathway has been implicated in motility in human trophoblast
cells~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}.

@@

The identification of multiple genes (\ac{CRH}, \ac{ADAM12},
\ac{KISS1}, \ac{PAPPA}, \ac{IL1RL1}, and \ac{PLAC1};
\cref{tab:human_lineage_specific}) which are associated with
pre-eclampsia which have increased expression on the \human lineage
may explain the larger proportion of pre-eclampsia observed in humans.
However, there are still genes which are associated with pre-eclampsia
which are also more highly expressed in the \mouse lineage such as
\ac{Afp} and \ac{Gjb2}. Furthermore, the onset of
pre-eclampsia is well before term; all of our placentas are term
placentas and provide limited information about the transcriptome
during the onset of pre-eclampsia.

@@

\subsubsection{Immune Regulation and Annexins}
The transcriptomes from multiple different species enabled us to
identify the set of expressed genes which are common to all placentas.
Components of the annexin complexes, including
\ac{ANXA2}, \ac{ANXA1}, \ac{S100A11}, and \ac{S100A10}
are expressed in all placentas examined, and argue for the evolution
importance of annexins in placenta function, where they likely
establish maternal-fetal
tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
We suggest that further research into molecular role of these annexin
complexes may illuminate how immune tolerance is maintained during
placentation.

@@

While Syncytin-1 and -2 are implicated in the formation of the
syncytiotrophoblast in
\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
of \ac{ERVFRD-1} and \ac{ERVW-1} are widely expressed in most
of the placentas analyzed here, they are not expressed in \bonobo,
\sheep, and \dog, even though these mammals form
syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
Because the formation of syncytiotrophoblast is the stem state of
Eutheria, Afrotheria, Euarchontoglires, and
Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
the gene(s) responsible for cytotrophoblast fusion to form
syncytiotrophoblast is likely to be highly expressed in the core
placenta. \ac{ANXA5} is known to be expressed at multiple stages
of placentation, and is capable of repairing membrane defects and
promoting the fusion of
cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
the presence of Ca$^{2+}$. While there are likely other genes which
are responsible for the positioning of cytotrophoblast cells in close
proximity to start the fusion process, we hypothesize that \ac{ANXA5} is
essential for the completion of fusion and the maintenance of membrane
repair during the massive flux of nutrients and waste in the placenta.

@@

The high expression of components of the EGFR signaling pathway in the
core placenta transcriptome (including \ac{GRB2}, \ac{RALB})
as well as annexins which are involved in the location of EGF receptor
localization (\ac{ANXA1} and \ac{ANXA2}) in the core placenta
is consistent with previous findings showing that alterations to the
EGF signaling system change invasion of trophoblast cells into the
spiral arteries~\cite{Armant.ea2015:Reducedexpressionofepidermal}.
Another component of the core placenta transcriptome, \ac{CD9},
has been shown to suppress \human extravillous trophoblast
invasion~\cite{Matsumoto.ea2016:CD9suppresseshumanextravillous}, and
we hypothesize that it is responsible for maintaining the balance
between invasion and attachment in the other placentas examined here
as well.

@@

The core placenta transcriptome contains genes which are involved in
maternal-fetal tolerance, such as the annexins, and genes responsible
for cell-cell and cell-matrix interactions, such as \ac{LGALS1}.
These genes were almost certainly present in the placenta of the
common mammalian ancestor. We hypothesize that modifications to their
expression or function will compromise the development and
organization of the placenta.

We also identified genes differentially expressed on the \human
lineage which are implicated in pre-eclampsia, including
\ac{ADAM12}, \ac{CRH}, and \ac{KISS1}.
